---
title: "Agent Skills"
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

shinyreact ships two [Agent Skills](https://agentskills.io/) so a coding agent
can build a shinyreact app without you pasting documentation into it:

- **`shinyreact-build-app`** builds a `ui.tsx`-pattern app from scratch.
- **`shinyreact-convert-app`** ports an existing Shiny app to the pattern. It
  inspects the source, drives the running app in a browser, describes it in
  plain English, then rewrites it against that description.

## Using the skills

Agents using [btw](https://posit-dev.github.io/btw/) discover the skills
automatically once shinyreact is attached.

To copy them into a project for Claude Code and other skill-aware tools:

```{r}
btw::btw_skill_install_package("shinyreact")
```

The installed files live at `system.file("skills", package = "shinyreact")`.

## Starter prompt

The skill doubles as the starter template: it knows the Vite setup, the file
layout, and the hooks, so you do not need to read about build tools before your
first app. With the shinyreact skills installed, paste this into your agent:

```
Using the `/shinyreact-build-app` skill, scaffold a new shinyreact app in this
directory with an R server. Set up the Vite build (src/ui.tsx compiled to
www/ui.js) and a hello-world UI: a text input whose value the server greets by
name. Run the build and start the app.
```

To port an existing app instead, point the agent at it: "Using the
`/shinyreact-convert-app` skill, convert the Shiny app in `./app.R` to the
shinyreact `ui.tsx` pattern."

## Python

The Python package ships the same two skills. Install them into a project with
[`library-skills`](https://github.com/tiangolo/library-skills):

```bash
uvx library-skills --claude   # Claude Code (.claude/skills)
uvx library-skills            # standard location (.agents/skills)
```
