---
title: "Testing wire payloads"
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

In a shinyreact app the contract between server and client is the JSON that
crosses the Shiny websocket: the values `reactive_output()` delivers, the
payloads `send_message()` pushes, and the values `useShinyInput()` sends back.
`wire_tap()` records those payloads in a
[shinytest2](https://rstudio.github.io/shinytest2/) test so you can assert on
them directly, without inspecting the rendered DOM.

## Setup

`wire_tap()` needs the shinytest2 package. Start the `AppDriver` with
`shiny.trace = TRUE` so every websocket frame is recorded in the app's logs:

```{r}
test_that("dist_data bins the waiting column", {
  app <- shinytest2::AppDriver$new(
    app_dir,
    options = list(shiny.trace = TRUE)
  )
  withr::defer(app$stop())

  tap <- shinyreact::wire_tap(app)
  tap$expect_input_value("bins", 30L)
  tap$expect_output_value("dist_data", function(d) d$breaks[[1]] == 43)
})
```

## Matchers

Each `expect_*` function takes a matcher and retries until it matches or a
timeout (10 seconds by default) elapses:

- A **function** is satisfied by a truthy return value. A function that errors
  on a payload's shape counts as a non-match, not a test failure.
- Any **other object** is compared with `identical()`. JSON numbers parse to
  integer when they have no fraction, so compare against `30L`, not `30`.

There is one `expect_*` per channel:

| Function | Channel |
|---|---|
| `expect_output_value(id, matcher)` | values delivered for `output$id` |
| `expect_message(id, matcher)` | `send_message(session, id, ...)` payloads |
| `expect_input_value(id, matcher)` | values the client sent for `input$id` |

Successive expectations on one channel assert an ordered subsequence: each
scans from just past the previous match, so a value that arrives between two
checks is never missed.

## Full histories

To inspect everything that crossed a channel, use the `all_*` functions:

```{r}
tap$all_output_values("dist_data")
tap$all_messages("notify")
tap$all_input_values("bins")
```

Input ids match the bare id or any `id:type` wire id, so use the id you wrote
in `useShinyInput()`.

## Python counterpart

`shinyreact.playwright.WireTap` in the Python package has the same methods and
semantics for Playwright tests. One divergence: `jsonlite::fromJSON()` maps a
JSON `null` output value to `NULL`, indistinguishable from an absent key, so
early `null` frames are dropped in R where Python records `None`.
