---
title: "Testing"
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

In a shinyreact app the contract between server and client is the JSON that
crosses the Shiny websocket: the values `reactive_output()` delivers, the
payloads `send_message()` pushes, and the values `useShinyInput()` sends back.
Both halves of this article assert that JSON — the first with no browser at
all, the second inside a real one.

## Testing the server with `testServer()`

A `ui.tsx` server contains only reactive computation, so "this input produces
that output value" *is* the server. `reactive_output()` is an ordinary render
function, so `shiny::testServer()` drives it with no browser and no client, and
`output$id` is the JSON value itself — no spec wrapper, no coercion:

```{r}
test_that("dist_data bins the waiting column", {
  shiny::testServer(app_dir, {
    session$setInputs(bins = 9)

    expect_named(output$dist_data, c("breaks", "counts"))
    expect_equal(
      unclass(output$dist_data$counts),
      c(16L, 37L, 30L, 16L, 14L, 57L, 67L, 29L, 6L)
    )
    expect_equal(output$dist_caption, "272 eruptions in 9 bins")
  })
})
```

`unclass()` is there because the app wraps its vectors in `I()` so a one-bin
result serializes as `[272]` rather than `272` — the `AsIs` class rides along
on the value the test sees.

Pass a directory containing the app, or the `server` function itself. Module
ids are namespaced as the session sees them (``output$`counter-label` ``), or
reachable through `session$makeScope("counter")`; a module server can also be
driven on its own:

```{r}
shiny::testServer(card_server, args = list(id = "left"), {
  session$setInputs(n = 7)
  expect_equal(output$label, "n=7")
})
```

`testServer()` **raises** rather than reporting a status. Reading an output
whose render function failed `req()` throws a silent error, which is how you
assert that an output produced nothing:

```{r}
expect_error(output$answer, class = "shiny.silent.error")
```

`examples/01-hello` and `examples/07-plotly` both carry a
`tests/testthat/test-outputs.R` written this way, runnable from the app
directory with `shiny::runTests()`.

The Python counterpart is `shiny.testserver.test_server()`, which reports
`.status` (`"ok"`, `"error"`, `"silent"`) instead of raising.

## Testing wire payloads with `wire_tap()`

`wire_tap()` records the payloads crossing the websocket in a
[shinytest2](https://rstudio.github.io/shinytest2/) test so you can assert on
them directly, without inspecting the rendered DOM. Reserve it for what
`testServer()` structurally cannot see: the values the *client* chooses to
send, real bindings, and real rendering.

### Setup

`wire_tap()` needs the shinytest2 package. Start the `AppDriver` with
`shiny.trace = TRUE` so every websocket frame is recorded in the app's logs:

```{r}
test_that("dist_data bins the waiting column", {
  app <- shinytest2::AppDriver$new(
    app_dir,
    options = list(shiny.trace = TRUE)
  )
  withr::defer(app$stop())

  tap <- shinyreact::wire_tap(app)
  tap$expect_input_value("bins", 30L)
  tap$expect_output_value("dist_data", function(d) d$breaks[[1]] == 43)
})
```

### Matchers

Each `expect_*` function takes a matcher and retries until it matches or a
timeout (10 seconds by default) elapses:

- A **function** is satisfied by a truthy return value. A function that errors
  on a payload's shape counts as a non-match, not a test failure.
- Any **other object** is compared with `identical()`. JSON numbers parse to
  integer when they have no fraction, so compare against `30L`, not `30`.

There is one `expect_*` per channel:

| Function | Channel |
|---|---|
| `expect_output_value(id, matcher)` | values delivered for `output$id` |
| `expect_message(id, matcher)` | `send_message(session, id, ...)` payloads |
| `expect_input_value(id, matcher)` | values the client sent for `input$id` |

Successive expectations on one channel assert an ordered subsequence: each
scans from just past the previous match, so a value that arrives between two
checks is never missed.

### Full histories

To inspect everything that crossed a channel, use the `all_*` functions:

```{r}
tap$all_output_values("dist_data")
tap$all_messages("notify")
tap$all_input_values("bins")
```

Input ids match the bare id or any `id:type` wire id, so use the id you wrote
in `useShinyInput()`.

### Python counterpart

`shinyreact.playwright.WireTap` in the Python package has the same methods and
semantics for Playwright tests. One divergence: `jsonlite::fromJSON()` maps a
JSON `null` output value to `NULL`, indistinguishable from an absent key, so
early `null` frames are dropped in R where Python records `None`.
